Fast Mini-Batch MCMC for Big Data Parameter Optimization

### 2016-05-13 nips_2016.tex
Updated abstract of paper

\title{Fast Mini-Batch MCMC for Big Data Bayesian Posterior Inference}

@@

\begin{document}
% \nipsfinalcopy is no longer used

@@

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have become one of the most popular techniques in machine
learning, but they can be intractably slow when doing Bayesian posterior estimation due to the need
to use information from all training data points during each sampling iteration. We propose a
mini-batch approach that uses information from a small subset of samples per iteration. This, along
with a new Metropolis-Hastings test, will allow us to sample faster to obtain higher-quality
parameter estimates. We implement our sampler in BIDMach and benchmark it with the state of the art.
{\color{blue} Daniel: I hope we can do this.}
{\color{blue} Xinlei: Compare with "cutting MH test budget" paper?}
\end{abstract}

\section{Introduction}

MCMC has become one of the most popular techniques in machine learning...

In this paper, we resolve these challenges\footnote{{\color{blue} Daniel: we can write these
``challenges'' later once the project becomes clearer}.} by proposing an MCMC algorithm for
Bayesian parameter estimation in which proposals are fast and mini-batch based. Our M-H test
acceptance probability is about 50 percent, reasonably high, which means our samples mix well.We
experimentally show that our approach is better than the state of the art.

{\color{blue} Daniel: we can fill out this section later.}

\section{Related Work}

@@

>>>>

\title{Fast Mini-Batch MCMC for Big Data Parameter Optimization}

@@

\begin{document}

\begin{comment}
\end{comment}

@@

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods are one of the most popular techniques in machine learning
for Bayesian inference. With large datasets, however, MCMC methods can be intractably slow due to
the need to compute the likelihood of all the data items. Borrowing from the optimization
literature, one intuitive way to make MCMC faster is to compute necessary gradients or likelihood
ratios based only on a random subset of the data each iteration. Such stochastic MCMC has received
increased research attention in recent years, but it is relatively difficult to develop formal
guarantees for those algorithms. Some stochastic MCMC methods also involve adaptively increasing the
minibatch size, which can nullify the benefits of the method if too much data is used. In addition,
with large datasets, it is unclear if obtaining a full posterior distribution is necessary or better
than deriving just a single MAP or point estimate of the parameters. In this paper, we present an
alternative perspective on stochastic MCMC based on deriving high-quality point estimates of
parameters in the presence of big data. To this end, we present a different Metropolis-Hastings test
designed to be more efficient in the state space. We also augment the target posterior distribution
with a temperature parameter, alleviating the need to adjust minibatch sizes.  We theoretically
show the benefits of our approach, and verify with empirical benchmarks on synthetic data and neural
network optimization.
\end{abstract}

\section{Introduction}\label{sec:introduction}

Markov chain Monte Carlo (MCMC) sampling is one of the most popular techniques for Bayesian
computation. In general, an MCMC algorithm will propose samples from a proposal distribution $q$,
and decide whether to accept or reject the new sample based on a Metropolis-Hastings
test~\cite{Metropolis1953,hastings70}. With traditional Bayesian posterior inference, the posterior
probability $p(\theta \mid x_1, \ldots x_N) \propto p(\theta)\prod_{i=1}^np(x_i \mid \theta)$ based
on i.i.d. data $\{x_i\}_{i=1}^n$ is the target distribution which one aims to approximate by
sampling. This formula is problematic when $n$ is large (e.g., on the order of billions) because the
likelihood must be evaluated based on each individual point.

and~\cite{RobbinsMonro1951}.

In this paper, we resolve these challenges by proposing an MCMC algorithm for Bayesian parameter
estimation in which proposals are fast and mini-batch based. Our M-H test acceptance probability is
about 50 percent, reasonably high, which means our samples mix well.We experimentally show that our
approach is better than the state of the art.

Our goal is using MH testing is in the context of MCMC model optimization, especially for neural
networks. We are also interested in using it for other big data problems. We are not interested in
exact inference of the posterior model parameters for a very large dataset.

The example you have been using, the mixture of two gaussians, illustrates this point well. The
"full data" posterior  based on 300 points has two sharp modes. The "cutting the MH budget" paper
takes on average about half the data points to make steps on this distribution. That's intuitive.
But it is also not very useful. The more data you have, the sharper the modes based on all the data.
On a non-toy problem (e.g. > 10,000 points), the posterior would collapse to two points from which
the MCMC simulation could likely never escape. And the "cutting the budget" method would still
require almost all the data to make steps.

For any real big-data problem (billions of points) the story would be the same.

We should argue instead that for model parameter inference, the goal is either MAP or ML estimation.
In both cases the goal is to find a good posterior mode. But the exact form of the distribution is
arbitrary, and e.g. non-unit powers of the distribution are fine. That corresponds to changing the
temperature of the distribution, or changing the scale factor in log delta space (which the factor
N/n is doing).

So rather than fixing the posterior temperature and trying to use adjustable mini-batch size to
achieve it (which ends up reducing to nearly-full-batch updates) we are proposing to run at the
natural temperature induced by a given minibatch size. We can always increase the mini-batch size to
reduce the temperature. But there is no need to do this beyond the point where the posterior mode is
sharp and well-defined. That will occur at some fixed number of points (e.g. a few thousand,
depending on the problem), and far less than the full dataset.

\section{Related Work}\label{sec:related_work}

@@

\subsection{Experiment 3}
